---
title: "3. Claret-Bruno model"
author:
  - Daniel Sabanés Bové
  - Francois Mercier
date: last-modified
editor_options: 
  chunk_output_type: inline
format:
  html:
    code-fold: show
    html-math-method: mathjax
cache: true
---

This appendix shows how the Claret-Bruno model can be implemented in a Bayesian framework using the `brms` package in R.

## Setup and load data

{{< include _setup_and_load.qmd >}}

{{< include _load_data.qmd >}}

## Claret-Bruno model

In the Claret-Bruno model we have again the baseline SLD and the growth rate as in the Stein-Fojo model. Then in addition we have the inhibition response rate $\psi_{p}$ and the treatment resistance rate $\psi_{c}$. The model is then:

$$
y^{*}(t_{ij}) = \psi_{b_{0}i} \exp \left\{
  \psi_{k_{g}i} t_{ij} -
  \frac{\psi_{pi}}{\psi_{ci}} (1 - \exp(-\psi_{ci} t_{ij}))
\right\}
$$

for positive times $t_{ij}$. Again, if the time $t$ is negative, i.e. the treatment has not started yet, then it is reasonable to assume that the tumor cannot shrink yet. That is, we have then $\psi_{pi} = 0$. Therefore, the final model for the mean SLD is:

$$
y^{*}(t_{ij}) = 
\begin{cases} 
\psi_{b_{0}i} \exp(\psi_{k_{g}i} t_{ij}) & \text{if } t_{ij} < 0 \\
\psi_{b_{0}i} \exp \{
  \psi_{k_{g}i} t_{ij} -
  \frac{\psi_{pi}}{\psi_{ci}} (1 - \exp(-\psi_{ci} t_{ij}))
\} & \text{if } t_{ij} \geq 0 
\end{cases}
$$

For the new model parameters we can again use log-normal prior distributions.

## Fit model

We can now fit the model using `brms`. The structure is determined by the model formula:

```{r}
#| label: specify_gsf_model

formula <- bf(sld ~ ystar, nl = TRUE) +
  # Define the mean for the likelihood
  nlf(
    ystar ~ 
      int_step(year > 0) * 
        (b0 * exp(kg * year - (p / c) * (1 - exp(-c * year)))) +
      int_step(year <= 0) * 
        (b0 * exp(kg * year))
  ) +
  # Define the standard deviation (called sigma in brms) as a 
  # coefficient tau times the mean.
  # sigma = tau * ystar is modelled on the log scale though, therefore:
  nlf(sigma ~ log(tau) + log(ystar)) +
  lf(tau ~ 1) +
  # Define nonlinear parameter transformations
  nlf(b0 ~ exp(lb0)) +
  nlf(kg ~ exp(lkg)) +
  nlf(p ~ exp(lp)) +
  nlf(c ~ exp(lc)) +
  # Define random effect structure
  lf(lb0 ~ 1 + (1 | id)) + 
  lf(lkg ~ 1 + (1 | id)) +
  lf(lp ~ 1 + (1 | id)) + 
  lf(lc ~ 1 + (1 | id))

# Define the priors
priors <- c(
  prior(normal(log(65), 1), nlpar = "lb0"),
  prior(normal(log(1.04), 1), nlpar = "lkg"),
  prior(normal(log(0.5), 0.1), nlpar = "lp"),
  prior(normal(log(0.5), 0.1), nlpar = "lc"),
  prior(normal(0, 3), lb = 0, nlpar = "lb0", class = "sd"),
  prior(normal(0, 3), lb = 0, nlpar = "lkg", class = "sd"),
  prior(normal(0, 3), lb = 0, nlpar = "lp", class = "sd"),
  prior(normal(0, 3), lb = 0, nlpar = "lc", class = "sd"),
  prior(normal(0, 3), lb = 0, nlpar = "tau")
)

# Initial values to avoid problems at the beginning
n_patients <- nlevels(df$id)
inits <- list(
  b_lb0 = array(3.61),
  b_lkg = array(-1.33),
  b_lp = array(-0.69),
  b_lc = array(-0.69),
  sd_1 = array(0.58),
  sd_2 = array(1.6),
  sd_3 = array(0.994),
  sd_4 = array(0.1),
  b_tau = array(0.161),
  z_1 = matrix(0, nrow = 1, ncol = n_patients),
  z_2 = matrix(0, nrow = 1, ncol = n_patients),
  z_3 = matrix(0, nrow = 1, ncol = n_patients),
  z_4 = matrix(0, nrow = 1, ncol = n_patients)
)

# Fit the model
save_file <- here("session-tgi/cb1.RData")
if (file.exists(save_file)) {
  load(save_file)
} else {
  fit <- brm(
    formula = formula,
    data = df,
    prior = priors,
    family = gaussian(),
    init = rep(list(inits), CHAINS),
    chains = CHAINS, 
    iter = WARMUP + ITER, 
    warmup = WARMUP, 
    seed = BAYES.SEED,
    refresh = REFRESH
  )
  save(fit, file = save_file)
}

# Summarize the fit
summary(fit)
```

todo cont here

## Parameter estimates

```{r}
#| label: gsf_post_processing

post_df <- as_draws_df(fit)
head(names(post_df), 10)

post_df <- post_df |>
  mutate(
    b0 = exp(b_lb0_Intercept + sd_id__lb0_Intercept^2 / 2),
    ks = exp(b_lks_Intercept + sd_id__lks_Intercept^2 / 2),
    kg = exp(b_lkg_Intercept + sd_id__lkg_Intercept^2 / 2),
    phi = plogis(b_tphi_Intercept),
    omega_0 = sd_id__lb0_Intercept,
    omega_s = sd_id__lks_Intercept,
    omega_g = sd_id__lkg_Intercept,
    omega_phi = sd_id__tphi_Intercept,
    cv_0 = sqrt(exp(sd_id__lb0_Intercept^2) - 1),
    cv_s = sqrt(exp(sd_id__lks_Intercept^2) - 1),
    cv_g = sqrt(exp(sd_id__lkg_Intercept^2) - 1),
    sigma = b_tau_Intercept
  )
```

Let's first look at the population level parameters:

```{r}
#| label: gsf_pop_params

mcmc_trace(post_df, pars = c("b0", "ks", "kg", "phi", "sigma"))
mcmc_pairs(post_df, pars = c("b0", "ks", "kg", "phi", "sigma"))
```

The trace plots look good. The chains seem to have converged and the pairs plot shows no strong correlations between the parameters. Let's check the table:

```{r}
#| label: gsf_post_summary

post_sum <- post_df |>
  select(b0, ks, kg, phi, omega_0, omega_s, omega_g, omega_phi, sigma) |>
  summarize_draws() |>
  gt() |>
  fmt_number(decimals = 3)
post_sum
```

So $\phi$ is estimated around 0.5. The other parameters are similar to the previous Stein-Fojo model, but we see a larger $k_s$ e.g.

## Observation vs model fit

We can now compare the model fit to the observations. Let's do this for the first 20 patients again:

```{r}
#| label: gsf_model_fit

pt_subset <- as.character(1:20)
df_subset <- df |> 
  filter(id %in% pt_subset)

df_sim <- df_subset |> 
  data_grid(
    id = pt_subset, 
    year = seq_range(year, 101)
  ) |>
  add_epred_draws(fit) |>
  median_qi()

df_sim |>
  ggplot(aes(x = year, y = sld)) +
  facet_wrap(~ id) +
  geom_ribbon(
    aes(y = .epred, ymin = .lower, ymax = .upper), 
    alpha = 0.3, 
    fill = "deepskyblue"
  ) +
  geom_line(aes(y = .epred), color = "deepskyblue") +
  geom_point(data = df_subset, color = "tomato") +
  coord_cartesian(ylim = range(df_subset$sld)) +
  scale_fill_brewer(palette = "Greys") +
  labs(title = "GSF model fit")
```

This also looks good. The model seems to capture the data well.

## With `jmpost`

This model can also be fit with the `jmpost` package. The corresponding function is `LongitudinalClaretBruno`. The statistical model is specified in the vignette [here](https://genentech.github.io/jmpost/main/articles/statistical-specification.html#claret-bruno-model).