We will use the publicly published tumor size data from the OAK study, see [here](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1009822). In particular we are using the [S1 data set](https://doi.org/10.1371/journal.pcbi.1009822.s006), which is the fully anonymized data set used in the publication. For simplicity, we have copied the data set in this GitHub repository.

```{r}
file_path <- here("data/journal.pcbi.1009822.s006.xlsx")

read_one_sheet <- function(sheet) {
    read_excel(file_path, sheet = sheet) |> 
    clean_names() |> 
    mutate(
        id = factor(as.character(patient_anonmyized)),
        day = as.integer(treatment_day),
        year = day / 365.25,
        target_lesion_long_diam_mm = case_match(
            target_lesion_long_diam_mm,
            "TOO SMALL TO MEASURE" ~ "2",
            "NOT EVALUABLE" ~ NA_character_,
            .default = target_lesion_long_diam_mm
        ),
        sld = as.numeric(target_lesion_long_diam_mm),
        sld = ifelse(sld == 0, 2, sld),
        study = factor(gsub("^Study_(\\d+)_Arm_\\d+$", "\\1", study_arm)),
        arm = factor(gsub("^Study_\\d+_Arm_(\\d+)$", "\\1", study_arm))
    ) |> 
    select(id, year, sld, study, arm)
}

tumor_data <- excel_sheets(file_path) |> 
    map(read_one_sheet) |> 
    bind_rows()

head(tumor_data)
summary(tumor_data)
```

For simplicity, we will for now just use study 4 (this is the OAK study), and we rename the patient IDs:

```{r}
#| label: subset_oak

df <- tumor_data |> 
  filter(study == "4") |> 
  na.omit() |>
  droplevels() |> 
  mutate(id = factor(as.numeric(id)))
```

Here we have `r nlevels(df$id)` patients. It is always a good idea to make a plot of the data. Let's look at the first 20 patients e.g.:

```{r}
#| label: plot_data

df |> 
  filter(as.integer(id) <= 20) |>
  ggplot(aes(x = year, y = sld, group = id)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ id) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme(legend.position = "none")
```
